---
title: "HW3"
author: "Hannah Zmuda"
date: "11/18/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library,include=FALSE}
```

## Question 1
### Problem
Use Monte Carlo simulation to investigate whether the empirical Type I error rate of the t-test is approximately equal to the nominal significance level $\alpha$, when the sampled population is non-normal. The t-test is robust to mild departures from normality. Discuss the simulation results for the cases where the sampled population is (i) $\chi^2$(1), (ii) Uniform(0,2), and (iii) Exponential(rate=1). In each case, test H0 : $\mu$ = $\mu_0$ vs H0 : $\mu$ $\ne$ $\mu_0$ , where $\mu_0$ is the mean of $\chi^2$(1), Uniform(0,2), and Exponential(1), respectively.

### Solution
```{r Q1}
#Part a: looking at a type I error
#all follow a similar algorithm as shown on page 193 in SCRR
n <- 100 #number of replicates
a <- 0.05 #significance level alpha
muA <- mean(rchisq(n, df = 1)) #mu0 in part a
muB <- mean(runif(n, 0, 2)) #mu0 in part b
muC <- mean(rexp(n, rate = 1)) #mu0 in part c
#become alternatives in the estimate power of a test (b).

m <- 1000 #number of replicates
pA <- numeric(m)
pB <- numeric(m)
pC <- numeric(m)

for(i in 1:m){
  xA <- rchisq(n, df = 1) #sample dist part a
  xB <- runif(n, 0, 2) #sample dist part b
  xC <- rexp(n, rate = 1) #sample dist part c
  
  ttestA <- t.test(xA, alternative = "two.sided",mu = muA)
  ttestB <- t.test(xB, alternative = "two.sided",mu = muB)
  ttestC <- t.test(xC, alternative = "two.sided",mu = muC)
  
  pA[i] <- ttestA$p.value
  pB[i] <- ttestB$p.value
  pC[i] <- ttestC$p.value
}

pHatA <- mean(pA < a)
pHatB <- mean(pB < a)
pHatC <- mean(pC < a)

seHatA <- sqrt(pHatA * (1- pHatA)/m)
seHatB <- sqrt(pHatB * (1- pHatB)/m)
seHatC <- sqrt(pHatC * (1- pHatC)/m)

cat("The means are: ", c(pHatA, pHatB, pHatC))
cat("The standard errors are: ", c(seHatA, seHatB, seHatC))

#Part b: Estimating power of a test and outputting empirical power curves of the t-test from the three previous situations




```


## Question 2



## Question 3



## Question 4



## Question 5

